---
title: "Multiple imputation versus debiased machine learning for a missing covariate"
author: "Dominic Obeng Koranteng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/", dpi = 200, fig.retina = 1)
library(ggplot2)
source("R/dgm.R")
source("R/performance.R")

sim <- readRDS("results/sim_results.rds")
perf <- summarise_performance(sim$results)
n_sim <- sim$settings$n_sim

method_labels <- c(full = "Full data (benchmark)", cc = "Complete cases",
                   mean_imp = "Mean imputation", mice = "MICE (pmm)",
                   smcfcs = "smcfcs", dml = "Debiased ML (AIPW)")
term_labels <- c("x" = "beta[1]~(x)", "x:z" = "beta[3]~(x %*% z)")
perf <- subset(perf, term %in% names(term_labels))
perf$method_f <- factor(method_labels[perf$method], levels = rev(method_labels))
perf$term_f <- factor(term_labels[perf$term], levels = term_labels)
perf$mech_f <- factor(perf$mechanism, levels = c("MCAR", "MAR", "MNAR"))
perf$rate_f <- factor(sprintf("%.0f%% missing", 100 * perf$rate))

# Look up one performance number, e.g. pv("MAR", 0.2, "cc", "x", "bias")
pv <- function(mech, rate, method, term, what) {
  perf[perf$mechanism == mech & perf$rate == rate & perf$method == method &
         perf$term == term, what]
}
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
pct <- function(x) paste0(formatC(100 * x, format = "f", digits = 1), "%")

# Shared look for the figures
ink <- "#0b0b0b"; ink2 <- "#52514e"; grid_col <- "#e6e5e0"
rate_cols <- c("20% missing" = "#2a78d6", "40% missing" = "#eb6834")
theme_study <- theme_minimal(base_size = 11) +
  theme(text = element_text(colour = ink), axis.text = element_text(colour = ink2),
        strip.text = element_text(colour = ink, face = "bold"),
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        panel.grid.major.x = element_line(colour = grid_col, linewidth = 0.3),
        legend.position = "top", legend.title = element_blank(),
        panel.spacing.x = unit(1.6, "lines"), panel.spacing.y = unit(1, "lines"),
        plot.background = element_rect(fill = "white", colour = NA))
```

## Aim

When a covariate is missing and the analysis model contains an interaction,
standard multiple imputation can bias the estimates. Two remedies exist:
imputation that respects the analysis model (smcfcs), and a doubly robust
estimator with machine-learned nuisance functions (debiased machine learning,
DML). This study compares both with complete-case analysis, mean imputation
and default MICE under MCAR, MAR and MNAR mechanisms. The design follows the
ADEMP structure of Morris, White and Crowther (2019).

## Design

**Data-generating mechanism.** Each dataset has *n* = `r sim$settings$n` units.

- $Z \sim N(0, 1)$ and $X = 0.5Z + \sqrt{0.75}\,E$ with $E \sim N(0, 1)$, so $X \sim N(0, 1)$ and $\mathrm{corr}(X, Z) = 0.5$.
- $Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ + U$, with $U \sim N(0, 1)$ and $(\beta_0, \beta_1, \beta_2, \beta_3) = (0, 1, 1, 0.5)$.

$X$ is incomplete; $Y$ and $Z$ are always observed. The probability that $X$
is missing is

| Mechanism | $P(X \text{ missing})$ | Depends on |
|---|---|---|
| MCAR | constant | nothing |
| MAR  | $0.05 + 0.75\,\mathrm{expit}(a_0 + Y + 0.5Z)$ | the outcome and $Z$ |
| MNAR | $0.05 + 0.75\,\mathrm{expit}(a_0 + 1.5X + 0.5Z)$ | $X$ itself and $Z$ |

Root-finding sets the intercept $a_0$ so that `r 100 * sim$settings$rates[1]`%
or `r 100 * sim$settings$rates[2]`% of $X$ is missing on average. The floor and
ceiling keep every unit's chance of being observed at 20% or more, so the
positivity condition that weighting needs holds by design.

**Estimands.** $\beta_1$, the coefficient of $X$, and $\beta_3$, the
$X \times Z$ interaction, in the linear model `y ~ x + z + x:z`.

**Methods.**

| Method | What it does |
|---|---|
| Full data | Fits the model before any values are deleted. A benchmark only. |
| Complete cases | Drops units with missing $X$. |
| Mean imputation | Replaces missing $X$ with the observed mean. |
| MICE (pmm) | `mice` default: predictive mean matching of $X$ on $Y$ and $Z$, then the interaction is formed from the imputed $X$. The imputation model omits the interaction, so it is incompatible with the analysis model. `r sim$settings$m` imputations, Rubin's rules. |
| smcfcs | Substantive-model-compatible imputation (Bartlett et al. 2015) with the analysis model supplied. `r sim$settings$m` imputations, Rubin's rules. |
| Debiased ML (AIPW) | Augmented inverse-probability-weighted estimating equation for regression with a missing covariate (Robins, Rotnitzky and Zhao 1994), in one-step form. A probability forest estimates $P(R = 1 \mid Y, Z)$, bounded below at 0.05. A two-learner super learner (random forest plus quadratic regression, weights chosen by out-of-sample error) estimates $E[X \mid Y, Z]$ and $\mathrm{Var}(X \mid Y, Z)$. 5-fold cross-fitting; sandwich standard errors. |

**How the DML settings were chosen.** I fixed the learners in development runs
that used different seeds from this study. Forests alone gave a stable but
upward-biased estimator, and adding the quadratic regression reduced the bias.
Two 200-replicate pilots on the main-study seeds then showed that the AIPW
equation can fail. When an observed unit with an extreme $X$ gets
a large weight, the equation's matrix comes close to singular. With response
probabilities bounded at 0.01, about 7% of datasets at 40% MAR gave inflated
standard errors. Raising the bound to 0.05 (weights at most 20) cut that to 4%,
but one estimate still missed by more than 40. I then switched to the one-step
form, which starts from a plug-in estimate and adds one AIPW correction step. The one-step form
never inverts that matrix and has the same large-sample behaviour. The results
below come from the revised estimator on those same pilot seeds, so treat the
DML rows as provisional. Replicates 201 to 1000 of the full run give an
untouched check.

I tried a quadratic outcome model (`y ~ x + I(x^2) + z`) first and dropped it:
its estimating equation needs $E[X^4 \mid Y, Z]$, and none of the learners I
tried gave stable estimates at this sample size.

MICE, smcfcs and DML all assume that $X$ is missing at random given $Y$ and
$Z$. Complete-case analysis assumes instead that missingness does not depend on
$Y$ once $X$ and $Z$ are known. Under the MNAR mechanism here that assumption
holds, so complete-case analysis should stay unbiased while the other three
should not.

**Performance measures.** Bias, empirical and model-based standard errors,
root mean squared error and the coverage of nominal 95% confidence intervals,
each with its Monte Carlo standard error (MCSE). Each scenario has
`r n_sim` replicates.

## Results

```{r bias, fig.width = 9, fig.height = 5.2, fig.cap = ""}
plot_dat <- subset(perf, method != "mean_imp")
ggplot(plot_dat, aes(x = bias, y = method_f, colour = rate_f)) +
  geom_vline(xintercept = 0, colour = ink2, linewidth = 0.4) +
  geom_errorbarh(aes(xmin = bias - 1.96 * bias_mcse, xmax = bias + 1.96 * bias_mcse),
                 height = 0, linewidth = 0.6, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(term_f ~ mech_f, labeller = labeller(term_f = label_parsed)) +
  scale_colour_manual(values = rate_cols) +
  labs(x = "Bias (estimate minus truth), with 95% Monte Carlo interval", y = NULL) +
  theme_study
```

*Figure 1. Bias of each method. Mean imputation is left out because its bias
is far off this scale (between `r f2(min(subset(perf, method == "mean_imp")$bias))`
and `r f2(max(subset(perf, method == "mean_imp")$bias))`); Tables 1 and 2 list it.*

```{r coverage, fig.width = 9, fig.height = 5.2, fig.cap = ""}
mc_band <- 1.96 * sqrt(0.95 * 0.05 / n_sim)
ggplot(plot_dat, aes(x = 100 * coverage, y = method_f, colour = rate_f)) +
  annotate("rect", xmin = 100 * (0.95 - mc_band), xmax = 100 * (0.95 + mc_band),
           ymin = -Inf, ymax = Inf, fill = grid_col, alpha = 0.8) +
  geom_vline(xintercept = 95, colour = ink2, linewidth = 0.4) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_grid(term_f ~ mech_f, labeller = labeller(term_f = label_parsed)) +
  scale_colour_manual(values = rate_cols) +
  scale_x_continuous(limits = c(floor(min(100 * plot_dat$coverage) / 10) * 10, 100),
                     breaks = seq(0, 100, 10)) +
  labs(x = "Coverage of 95% confidence intervals (%)", y = NULL) +
  theme_study
```

*Figure 2. Coverage of nominal 95% intervals. The grey band shows the range
expected from Monte Carlo error alone if true coverage were 95%.*

```{r tables, results = "asis"}
cell <- function(est, mcse, digits = 3) {
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), est, mcse)
}
scen_order <- c("MCAR_20", "MAR_20", "MNAR_20", "MCAR_40", "MAR_40", "MNAR_40")
make_table <- function(term, what, mcse, digits, scale = 1) {
  d <- perf[perf$term == term, ]
  tab <- sapply(scen_order, function(s) {
    dd <- d[d$scenario == s, ]
    dd <- dd[match(names(method_labels), dd$method), ]
    cell(scale * dd[[what]], scale * dd[[mcse]], digits)
  })
  tab <- cbind(Method = method_labels, tab)
  colnames(tab) <- c("Method", "MCAR 20%", "MAR 20%", "MNAR 20%", "MCAR 40%", "MAR 40%", "MNAR 40%")
  tab
}
cat("**Table 1. Bias (MCSE).**\n\n*Coefficient of x*\n")
print(knitr::kable(make_table("x", "bias", "bias_mcse", 3), row.names = FALSE))
cat("\n*Interaction x:z*\n")
print(knitr::kable(make_table("x:z", "bias", "bias_mcse", 3), row.names = FALSE))
cat("\n**Table 2. Coverage of 95% intervals, % (MCSE).**\n\n*Coefficient of x*\n")
print(knitr::kable(make_table("x", "coverage", "coverage_mcse", 1, 100), row.names = FALSE))
cat("\n*Interaction x:z*\n")
print(knitr::kable(make_table("x:z", "coverage", "coverage_mcse", 1, 100), row.names = FALSE))
cat("\n**Table 3. Empirical SE / model-based SE / RMSE for the interaction x:z.**\n\n")
d <- perf[perf$term == "x:z", ]
tab3 <- sapply(scen_order, function(s) {
  dd <- d[d$scenario == s, ]
  dd <- dd[match(names(method_labels), dd$method), ]
  sprintf("%.3f / %.3f / %.3f", dd$emp_se, dd$mod_se, dd$rmse)
})
tab3 <- cbind(Method = method_labels, tab3)
colnames(tab3) <- c("Method", "MCAR 20%", "MAR 20%", "MNAR 20%", "MCAR 40%", "MAR 40%", "MNAR 40%")
print(knitr::kable(tab3, row.names = FALSE))
```

```{r timing}
secs <- aggregate(seconds ~ method, data = unique(sim$results[, c("scenario", "rep", "method", "seconds")]), FUN = median)
secs$method <- method_labels[secs$method]
secs <- secs[match(method_labels, secs$method), ]
```

Median computing time per dataset: MICE `r f2(secs$seconds[secs$method == "MICE (pmm)"])` s,
smcfcs `r f2(secs$seconds[secs$method == "smcfcs"])` s,
debiased ML `r f2(secs$seconds[secs$method == "Debiased ML (AIPW)"])` s.

## Findings

```{r findings, child = "findings.Rmd"}
```

## Reproducibility

Run `Rscript run_simulation.R` from the project folder, then knit this file.
Replicate *r* in every scenario generates its data from seed
`r sim$settings$base_seed` + *r* (Mersenne-Twister), and method *j* then runs
from seed `r sim$settings$base_seed` + *j* × 10^6 + *r*. You can regenerate
any single dataset or rerun any single method, and the results do not depend
on the number of parallel workers.

<details>
<summary>Session information for this run</summary>

```{r session}
sim$session
```

</details>

## References

- Bartlett JW, Seaman SR, White IR, Carpenter JR (2015). Multiple imputation of covariates by fully conditional specification: accommodating the substantive model. *Statistical Methods in Medical Research* 24(4):462-487.
- Chernozhukov V, Chetverikov D, Demirer M, Duflo E, Hansen C, Newey W, Robins J (2018). Double/debiased machine learning for treatment and structural parameters. *The Econometrics Journal* 21(1):C1-C68.
- Morris TP, White IR, Crowther MJ (2019). Using simulation studies to evaluate statistical methods. *Statistics in Medicine* 38(11):2074-2102.
- Robins JM, Rotnitzky A, Zhao LP (1994). Estimation of regression coefficients when some regressors are not always observed. *Journal of the American Statistical Association* 89(427):846-866.
- Seaman SR, Bartlett JW, White IR (2012). Multiple imputation of missing covariates with non-linear effects and interactions: an evaluation of statistical methods. *BMC Medical Research Methodology* 12:46.
- van Buuren S, Groothuis-Oudshoorn K (2011). mice: Multivariate imputation by chained equations in R. *Journal of Statistical Software* 45(3):1-67.
